The Galactic Midplane is Not a Plane: Implications for {\em Gaia}

### 2019-04-22 paper/ms.tex
revised abstract

@@

% \correspondingauthor{Angus Beane}
\email{abeane@sas.upenn.edu}

@@

Stellar actions, computed for stars from their six-dimensional phase space measurement and
an assumed Galactic potential, are used to label and distinguish orbits. In
principle, actions have the attractive quality of being invariant and thus
good labels of a star's orbit. Of course, inaccurate measurements of the 
phase space position and uncertainities in the assumed Galactic potential
induce an error in the computed action. We show that, in addition to these
complications, a systematic bias in the phase space position induces a
phase-dependence in the actions, which we interpret as an error. The induced error distribution is
non-Gaussian and bimodal, 
with neither of the modes peaking on the true action value. 
A midplane error of $\sim30\,\pc$ for a thin disk orbit or
$\sim225\,\pc$ for a thick disk orbit is all that is necessary to
induce a $100\,\%$ error in the vertical action $J_z$.
orbits, respectively. Furthermore, we show that the local midplane varies by
$\sim100\,\pc$ at the Solar circle in Milky Way-mass cosmological zoom-in
simulations from the FIRE collaboration. Thus, current state-of-the-art action
calculations --- which assume the global and local midplanes are the same ---
include a systematic offset in $z$. Variation in the local standard of rest
will induce similar issues. The variation of the midplane must be taken into
account when performing dynamical modelling across large regions of the disk
accessible with the {\em Gaia} mission.

>>>>

%mm substantive Mac Low changes marked so.  Minor changes done silently.

@@

% \correspondingauthor{Angus Beane}
%mm [will this address go away during the review process?.  Do you
%want to give a "current address" as a second footnote?]
\email{abeane@sas.upenn.edu}

@@

Stellar actions, computed for stars from their six-dimensional phase space measurement and
an assumed Galactic potential, are used to label and distinguish orbits. In
principle, actions have the attractive quality of being invariant and thus
good labels of a star's orbit. 
%mm Of course,
    Such analyses are often used to interpret data from the {\em Gaia}
    mission. However, 
inaccurate measurements of the 
phase space position and uncertainties in the assumed Galactic potential
induce an error in the computed action. We show that, in addition to these
complications, a systematic bias in the phase space position induces a
phase-dependence in the actions, which we interpret as 
%mm an 
    a systematic
error. The induced error distribution is
non-Gaussian and bimodal, 
with neither of the modes peaking on
%mm  the true action value. [the error in a Gaussian peaks on zero] 
    the null value.
A midplane error of $\sim30\,\pc$ for a thin disk orbit or
$\sim225\,\pc$ for a thick disk orbit is 
%mm all that is necessary 
      sufficient
to induce a $100\,\%$ error in the vertical action $J_z$.
%mm [fragment?] orbits, respectively. 
Furthermore, we show that the local midplane varies by
$\sim100\,\pc$ at the Solar circle in Milky Way-mass cosmological zoom-in
simulations from the FIRE collaboration. Thus, current state-of-the-art action
calculations --- which assume the global and local midplanes are the same ---
include a systematic 
%mm offset in $z$.
   vertical offset variation. 
Variation in the local standard of rest
%mm will induce 
     induces
similar issues. The variation of the midplane must be taken into
account when performing dynamical modelling across
the large regions of the disk
accessible 
%mm with the {\em Gaia} mission. [moved first mention up to top of
%abstract, given that this is in the title.]
    to {\em Gaia}.
